### 2020-07-18 paper.tex
Pass through section 7

\section{A DYNAMIC CHECK FOR DETECTING BUGS}

@@

\begin{algorithm}[h]
\begin{algorithmic}
\Procedure{imcmc-dynamic-check}{$p$, $q$, $f$}
    \State $x \sim p(\cdot)$ \Comment{Sample state from prior}
    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
    \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
    \State {\bf assert} $p(x')q_{x'}(y') > 0$ \Comment{Support check}
    \State {\bf assert} $f(x', y') = (x, y)$ \Comment{Involution check}
    \State {\bf assert} $|\text{Rd}^{\textit{cont}}_{f,x,y}| = |\text{Wr}^{\textit{cont}}_{f,x,y}|$ \Comment{Dimension check}
\EndProcedure
\end{algorithmic}
\caption{Dynamic Check for Bugs}
\label{alg:dynamic-check}
\end{algorithm}

To test the user's algorithm, Gen first samples a model trace $x$ from the model prior $p$, and auxiliary randomness $y$ from the kernel $q_x$. It then runs three tests:

@@

The reversible jump MCMC framework involves a set of `models' $h \in \mathcal{H}$, and a prior distribution on models $p(h)$.
For each model, there is a latent continuous parameter vector $\theta_h \in \mathbb{R}^{n(h)}$ where $n(h)$ is the dimension of model $h$, and a likelihood function $L_{D,h}(\theta_h)$ for each $h$ given data $D$.
The latent state $x$ is a pair $(h, \theta_h)$ of model and continuous parameter.
There is a set of \emph{move types} $\mathcal{M}$.
Each move type $m \in \mathcal{M}$ is associated with an unordered pair of models $(h_1, h_2)$ and a dimensionality $d(m)$ such that $d(m) \ge n(h_1)$ and $d(m) \ge n(h_2)$ (zero, one, or more than one move types may be associated with a given pair of models).
For each latent state $x = (h, \theta_h)$, there is a probability distribution $q_{x}(m)$ on move types such that $q_{x}(m) > 0$ implies that $h$ is one of the models for move type $m$.
For each move type $m \in \mathcal{M}$ between $h_1$ and $h_2$ there is a pair of continuously differentiable bijections $g_{m, h_1 \to h_2} : \mathbb{R}^{d(m)} \to \mathbb{R}^{d(m)}$ and $g_{m, h_2 \to h_1} := g_{m, h_1 \to h_2}^{-1}$, and a pair of proposal densities $q_{m,h_1 \to h_2}(u_{h_1 \to h_2})$ and $q_{m,h_2 \to h_1}(u_{h_2 \to h_1})$ where $u_{h_1 \to h_2} \in \mathbb{R}^{d(m) - n(h_1)}$ and $u_{h_2 \to h_1} \in \mathbb{R}^{d(m) - n(h_2)}$.
A proposal is made from state $x = (h, \theta_h)$ by (i) sampling a move type $m \sim q_{x}(\cdot)$, and (ii) sampling continuous variable $u \sim q_{m,h \to h'}(\cdot)$ for $(h, h')$ associated with $m$, and (iii) computing $(\theta_{h}', u') := g_{m,h \to h'}(\theta_h, u)$, and proposing new state $x' = (h', \theta_h')$.
The move is accepted with probability:
\begin{equation}
{
\min\left\{ 1,
\frac{p(h')}{p(h)} \frac{p_{h'}(\theta_{h'})}{p_h(\theta_h)} \frac{L_{D,h'}(\theta_h')}{L_{D,h}(\theta_h)}
\frac{q_{x'}(m)}{q_{x}(m)}
\frac{q_{m,h' \to h}(u)}{q_{m,h \to h'}(u)}
\right\}
}
\end{equation}

@@

State-dependent mixture proposal.
Depending on the current state of the model, a different mixture distribution is used for the proposal.
Implementing such a move requires computing the mixture probability under the forward move and under the reverse move, which may differ.

$q_x(m)$ -- the mixture distribution

$q_m(x')$ -- the proposal we are mixing over

The first part of the auxiliary probabilistic program $Q$ determines which proposal to apply by sampling from a mixture distribution.
Here, the program picks a random node in the tree, by doing a stochastic walk of the tree that terminates at the chosen node.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
prev_cov_function = trace[:cov_function]
path ~ walk_tree(prev_cov_function, ..)
\end{lstlisting}}
\end{tabular}
\end{center}
The first part of the involution copies the random choices corresponding to the mixture from the input auxiliary trace to the output auxiliary trace.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@copy(q_in, :path, q_out, :path)
\end{lstlisting}}
\end{tabular}
\end{center}

The code that walks the tree uses the following recursion, which results in a probability distribution that assigns exponentially lower probability to nodes that are deeper in the tree.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
if ({:done} ~ bernoulli(0.5))
  return path
elseif ({:recurse_left} ~ bernoulli(0.5))
  path = (path..., :left_node)
  return ({:left} ~ walk_tree(node.left, path))
else
  path = (path..., :right_node)
  return ({:right} ~ walk_tree(node.right, path))
end
\end{lstlisting}}
\end{tabular}
\end{center}
The resulting mixture distributions for two trees are:
\begin{center}
\includegraphics[width=0.8\linewidth]{figures/biased-tree-distribution.pdf} % TODO improve quality
\end{center}
Because the mixture is specified using a probabilistic program, it is straightforward to modify the program to define a different mixture distribution.
The code below specifies a mixture distribution that is uniform over all nodes in the tree.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
n1 = size(node.left); n2 = size(node.right)
if ({:done} ~ bernoulli(1 / (1 + n1 + n2)))
  return path
elseif ({:recurse_left} ~ bernoulli(n1 / (n1+n2))
  path = (path..., :left_node)
  return ({:left} ~ walk_tree(node.left, path))
else
  path = (path..., :right_node)
  return ({:right} ~ walk_tree(node.right, path))
end
\end{lstlisting}}
\end{tabular}
\end{center}
The resulting distributions are:
\begin{center}
\includegraphics[width=0.8\linewidth]{figures/unbiased-tree-distribution.pdf}
\end{center}
Note that in the first case, the ratio of mixture probabilities is either $1$, $0.5$, or $2$ depending on whether the previous and new subtrees are leaf or internal nodes.
In the second case, the ratio of mixture probabilities is the ratio of sizes of the two trees.
%Automating the computation of the mixture probabilities

The rest of the kernel specifies what happens for each chosen node.
The second part of the auxiliary probabilistic program $Q$ implements a common motif for Metropolis-Hastings algorithms, where the proposal samples new values for some random choices:
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
new_subtree ~ cov_function_prior()
\end{lstlisting}}
\end{tabular}
\end{center}
and the involution swaps the old value with the newly proposed value:
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@copy(p_in, subtree_kddress, q_out, :new_subtree)
@copy(q_in, :new_subtree, p_out, subtree_kddress)
\end{lstlisting}}
\end{tabular}
\end{center}
Here the values happen to be trees that themselves have a random number of random choices, but the same motif is also used in simpler applications of Metropolis-Hastings.

Here, the proposal we are mixing over is easy to implement -- it involvs resampling a subtree from the prior, so the Jacobian is 1 and there is cancellation between the model and the proposal.

But computing the mixture probabilities is complex, because the mixture is a mixture of either $n_1$ or $n_2$ elements, which depends on the size of the tree.

Show the mixture probabilities in math notation?
If both subtrees not leaf nodes, then the probabilities are the same.
If both subtrees are leaf nodes, then the probabilities are different.
If one subtree is a leaf node and the other is an internal node then the ratio is either $0.5$ or $2$ respectively.
\begin{equation}
q
\end{equation}

TODO: relate to `resimulation Metropolis-Hastings' in general-purpose probabilistic programs.

Of course, it is straightforward to replace the prior with a custom data-driven proposal distribution on subtrees.

>>>>

\section{DYNAMICALLY DETECTING BUGS IN INVOLUTIONS}

@@

\begin{algorithm}[h]
\begin{algorithmic}
\Procedure{involutive-mcmc-check}{$p$, $q$, $f$}
    \State $x \sim p(\cdot)$ \Comment{Sample state from prior}
    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
    \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
    \State {\bf assert} $p(x')q_{x'}(y') > 0$ \Comment{Support check}
    \State {\bf assert} $f(x', y') = (x, y)$ \Comment{Involution check}
    \State {\bf assert} $|\text{Rd}^{\textit{cont}}_{f,x,y}| = |\text{Wr}^{\textit{cont}}_{f,x,y}|$ \Comment{Dimension check}
\EndProcedure
\end{algorithmic}
\caption{Dynamic Check for Bugs}
\label{alg:dynamic-check}
\end{algorithm}

To test the user's algorithm, Gen first samples a model trace $x$ from the model prior $p$, and auxiliary trace $y$ from the $q_x$. It then runs three tests:

@@

The reversible jump MCMC framework involves a set of `models' $h \in \mathcal{H}$, and a prior distribution on models $p(h)$.
For each model, there is a latent continuous parameter vector $\theta_h \in \mathbb{R}^{n(h)}$ where $n(h)$ is the dimension of model $h$, and a likelihood function $L_{D,h}(\theta_h)$ for each $h$ given data $D$.
The latent state $x$ is a pair $(h, \theta_h)$ of model and continuous parameter.
There is a set of \emph{move types} $\mathcal{M}$.
Each move type $m \in \mathcal{M}$ is associated with an unordered pair of models $(h_1, h_2)$ and a dimensionality $d(m)$ such that $d(m) \ge n(h_1)$ and $d(m) \ge n(h_2)$ (zero, one, or more than one move types may be associated with a given pair of models).
For each latent state $x = (h, \theta_h)$, there is a probability distribution $q_{x}(m)$ on move types such that $q_{x}(m) > 0$ implies that $h$ is one of the models for move type $m$.
For each move type $m \in \mathcal{M}$ between $h_1$ and $h_2$ there is a pair of continuously differentiable bijections $g_{m, h_1 \to h_2} : \mathbb{R}^{d(m)} \to \mathbb{R}^{d(m)}$ and $g_{m, h_2 \to h_1} := g_{m, h_1 \to h_2}^{-1}$, and a pair of proposal densities $q_{m,h_1 \to h_2}(u_{h_1 \to h_2})$ and $q_{m,h_2 \to h_1}(u_{h_2 \to h_1})$ where $u_{h_1 \to h_2} \in \mathbb{R}^{d(m) - n(h_1)}$ and $u_{h_2 \to h_1} \in \mathbb{R}^{d(m) - n(h_2)}$.
A proposal is made from state $x = (h, \theta_h)$ by (i) sampling a move type $m \sim q_{x}(\cdot)$, and (ii) sampling continuous variable $u \sim q_{m,h \to h'}(\cdot)$ for $(h, h')$ associated with $m$, and (iii) computing $(\theta_{h}', u') := g_{m,h \to h'}(\theta_h, u)$, and proposing new state $x' = (h', \theta_h')$.
The move is accepted with probability:
\begin{equation*}
{\scriptstyle
\min\left\{ 1,
\frac{p(h')}{p(h)} \frac{p_{h'}(\theta_{h'})}{p_h(\theta_h)} \frac{L_{D,h'}(\theta_h')}{L_{D,h}(\theta_h)}
\frac{q_{x'}(m)}{q_{x}(m)}
\frac{q_{m,h' \to h}(u)}{q_{m,h \to h'}(u)}
|J g_{m, h\to h'}|(\theta_h, u)
\right\}
}
\end{equation*}

@@

Figure~\ref{fig:structure-learning} shows automated involutive MCMC being applied to fully Bayesian inference over the covariance function of a Gaussian process, where the prior on covariance functions (Figure~\ref{fig:structure-learning}a) is based on a probabilistic context-free grammar.
The inference algorithm is based on an involutive MCMC kernel that uses a state-dependent mixture of proposals.
A variant of this inference algorithm was previously studied in~\citep{schaechtle2016time,saad2019bayesian} based on a model of~\citet{grosse2012exploiting}.

At each iteration of the MCMC algorithm, the auxiliary probabilistic program $\mathcal{Q}$ (Figure~\ref{fig:structure-learning}d) first picks a random node in the parse tree of the covariance function, by doing a stochastic walk of the existing parse tree that terminates at the chosen node.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
prev_cov_function = trace[:cov_function]
path ~ walk_tree(prev_cov_function, ..)
\end{lstlisting}}
\end{tabular}
\end{center}
The first part of the involution (Figure~\ref{fig:structure-learning}e) copies the random choices made during this walk from the input auxiliary trace to the output auxiliary trace.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@copy(q_in, :path, q_out, :path)
\end{lstlisting}}
\end{tabular}
\end{center}
The code that walks the tree uses the following recursion, which results in a probability distribution that assigns exponentially lower probability to nodes that are deeper in the tree.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
if ({:done} ~ bernoulli(0.5))
  return path
elseif ({:recurse_left} ~ bernoulli(0.5))
  path = (path..., :left_node)
  return ({:left} ~ walk_tree(node.left, path))
else
  path = (path..., :right_node)
  return ({:right} ~ walk_tree(node.right, path))
end
\end{lstlisting}}
\end{tabular}
\end{center}
The resulting distributions on selected nodes for two example trees are shown below:
\begin{center}
\includegraphics[width=0.8\linewidth]{figures/biased-tree-distribution.pdf} % TODO improve quality
\end{center}
The other parts of the auxiliary probabilistic program and the involution program specifies a proposal distribution for the subtree of the parse tree that is rooted at the chosen node.
In particular, the rest of the auxiliary probabilistic program $\mathcal{Q}$ proposes a new subtree by sampling from the same process used to recursively define the prior distribution:
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
new_subtree ~ cov_function_prior()
\end{lstlisting}}
\end{tabular}
\end{center}
The involution program swaps the old subtree with the newly proposed subtree:
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@copy(p_in, subtree_kddress, q_out, :new_subtree)
@copy(q_in, :new_subtree, p_out, subtree_kddress)
\end{lstlisting}}
\end{tabular}
\end{center}
Note that the subtrees here happen to be trees that themselves have a random number of random choices, but the same general construction for a mixture of proposals below can also be used for simpler Metropolis-Hastings kernels:
\begin{enumerate}
\item The auxiliary probabilistic program samples from a distribution over different sets of random choices that will be proposed to
(in this case, each set is a different subtree of the parse tree).
\item The auxiliary probabilistic program them samples new values for those random choices (in this case, a new subtree).
\item The involution program swaps the previous values of those random choices with their new values, by swapping data between the model trace and the auxiliary trace.
\item The involution program copies the choice of what subset of random choices to propose to from the input auxiliary trace to the output auxiliary trace.
\end{enumerate}

Because the mixture distribution is specified using a probabilistic program, it is straightforward to modify the program to define a different mixture distribution.
The code below specifies a mixture distribution that is uniform over all nodes in the tree.
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
n1 = size(node.left); n2 = size(node.right)
if ({:done} ~ bernoulli(1 / (1 + n1 + n2)))
  return path
elseif ({:recurse_left} ~ bernoulli(n1 / (n1+n2))
  path = (path..., :left_node)
  return ({:left} ~ walk_tree(node.left, path))
else
  path = (path..., :right_node)
  return ({:right} ~ walk_tree(node.right, path))
end
\end{lstlisting}}
\end{tabular}
\end{center}
The resulting distributions, for two input trees, are:
\begin{center}
\includegraphics[width=0.8\linewidth]{figures/unbiased-tree-distribution.pdf}
\end{center}
Note that the mixture probabilities do not in general cancel, and must be accounted for in the acceptance probability calculation.
For the original mixture distribution, the ratio of mixture probabilities is either $1$, $0.5$, or $2$ depending on whether the previous and new subtrees are leaf or internal nodes.
For this alternative mixture distribution, the ratio of mixture probabilities is the ratio of sizes of the two trees.
In both cases, our automated involutive MCMC algorithm automatically computes the acceptance probability.
